Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
more

A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite : \Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \stypes{\Sigma}$.
We write $t\typ{\tau}$ to denote a metavariable ranging over terms of
type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in $\I$,
and each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$;
otherwise, it is \emph{$T$-unsatisfiable}.
A formula $\varphi$ entails $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the type and function symbols in this theory as \emph{uninterpreted}.

@@

For example, the set $\{ \forallf{\con{f}} \vec x.\; \con{f}( x ) \teq 0 \}$ is admissible.
A set of admissible function definitions can include definitions of non-terminating functions, e.g.
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible,
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ), \forallf{\con{g}}.\;\, \lite\bigl( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) \bigr) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}}.\;\, \lite( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) )$.

>>>>

Our setting is a monomorphic (or many-sorted) first-order logic, such as
the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function symbol $\lite : \Bool \times \tau \times \tau \rightarrow \tau$ (``if--then--else'')
for each $\tau \in \stypes{\Sigma}$.
We write $t\typ{\tau}$ to denote a metavariable ranging over terms of
type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I$,
and each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$;
otherwise, it is \emph{$T$-unsatisfiable}.
A formula $\varphi$ entails $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the type and function symbols in this theory as \emph{uninterpreted}.

@@

For example, the set $\{ \forallf{\con{f}} \vec x.\; \con{f}( x ) \teq 0 \}$ is admissible.
A set of admissible function definitions can include definitions of nonterminating functions, e.g.
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible,
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ), \forallf{\con{g}}.\;\, \lite\bigl( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) \bigr) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}}.\;\, \lite( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) )$.
